---
title: "PA1_template.Rmd"
author: John Broberg
output: md_document
---


```{r echo = TRUE, results= "hide"}
# Clear Variables
rm(list = ls())

# Attach packages
library(plyr)
library(dplyr)
library(lubridate)
library(ggplot2)

```

# Loading and preprocessing the data

### Show any code that is needed to

1. Load the data (i.e. read.csv())

```{r echo = TRUE}
# Create data folder to store files
if(!file.exists("./data")){dir.create("./data")}

# Download raw data
fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileURL, destfile="./data/activity.zip")

# Unzip file
unzip(zipfile="./data/activity.zip", exdir="./data")

# Read in file(s)
activity <- read.csv("./data/activity.csv")

# Review data
dim(activity)
head(activity)
tail(activity)
summary(activity)
names(activity)
str(activity)

```
2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r, echo = TRUE}
# Total steps by date, remove NA
total.steps <- tapply(activity$steps, activity$date, FUN = sum, na.rm = TRUE)

# Format date
activity$date <- ymd(activity$date)
```

# What is mean total number of steps taken per day?
```{r, echo = TRUE}
mean(total.steps)
```
*For this part of the assignment, you can ignore the missing values in the dataset.*

1. Calculate the total number of steps taken per day
```{r, echo = TRUE}
stepsPerDay <- activity %>%
    filter(!is.na(steps)) %>%
    group_by(date) %>%
    summarize(steps = sum(steps)) %>%
    print
```

2. Make a histogram of the total number of steps taken each day
```{r, echo = TRUE}
plot1 <- ggplot(stepsPerDay, aes(date, steps/1000))
plot1 + geom_bar(stat="identity") + 
    labs(x = "", y = "Steps (thousand)") +
    labs(title= "Total Number of Steps Taken Each Day")
```

3. Calculate and report the mean and median of the total number of steps taken per day
```{r, echo = TRUE}
mean(total.steps)
median(total.steps)
```

# What is the average daily activity pattern?
```{r, echo = TRUE}
stepsPerInterval <- activity %>%
    filter(!is.na(steps)) %>%
    group_by(interval) %>%
    summarize(steps = mean(steps)) %>%
    print
```

1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r, echo = TRUE}
plot(stepsPerInterval, type = "l", xlab = "Interval", ylab = "Average Steps")
title(main = "Average Steps Taken by Interval")
```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r, echo = TRUE}
stepsPerInterval[which.max(stepsPerInterval$steps),]$interval
```

# Imputing missing values

*Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.*

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r, echo = TRUE}
# Total number of missing values (i.e. NA) in activity
sum(is.na(activity))
```

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

        Using dplyr's mutate function, we'll change all NA values to the mean of all steps.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r, echo = TRUE}
# Create dataset with NA changed to mean of all steps
activityNAtoMean <- activity %>%
    group_by(interval) %>%
    mutate(steps = ifelse(is.na(steps), mean(steps, na.rm = TRUE), steps))
summary(activityNAtoMean)
```

4. Make a histogram of the total number of steps taken each day
```{r, echo = TRUE}
activityNAtoMean.steps <- activityNAtoMean %>%
    group_by(date) %>%
    summarize(steps = sum(steps))
    
ggplot(activityNAtoMean.steps, aes(x=date, y=steps)) +
    geom_bar(stat = "identity") +
    labs(x = "", y = "Steps (Imputed)", title = "Total Steps by Date (Imputed)")
```

- ...and Calculate and report the mean and median total number of steps taken per day. 
```{r, echo = TRUE}
imputed.steps <- tapply(activityNAtoMean$steps, activityNAtoMean$date, FUN = sum, na.rm = TRUE)
activityNAtoMean$date <- ymd(activityNAtoMean$date)
mean(imputed.steps)
median(imputed.steps)
```

- Do these values differ from the estimates from the first part of the assignment? 
```{r, echo = TRUE}
if(mean(total.steps)==mean(imputed.steps)) {
    "No, mean values are equal."
}   else {
    "Yes, mean values differ."
}

if(median(total.steps)==median(imputed.steps)) {
    "No, median values are equal." 
}   else {
    "Yes, median values differ."
}
```

- What is the impact of imputing missing data on the estimates of the total daily number of steps?

Imputing the missing data impacted the estimates by the following differences:
```{r, echo = TRUE}
summary(imputed.steps) - summary(total.steps)
```

# Are there differences in activity patterns between weekdays and weekends?

*For this part the weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.*

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r, echo = TRUE}
weekdayWeekend <- function(date){
    if(weekdays(as.Date(date)) %in% c("Saturday", "Sunday")) {
        "weekend"
    }else {
        "weekday"
    }
}

activity$daytype <- as.factor(sapply(activity$date, weekdayWeekend))
summary(activity)
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.
```{r, echo = TRUE}
par(mfrow = c(2,1))

stepsWeekday <- aggregate(steps ~ interval, data = activity, 
    subset = activity$daytype == "weekday", FUN = mean)
stepsWeekend <- aggregate(steps ~ interval, data = activity, 
    subset = activity$daytype == "weekend", FUN = mean)

plot(stepsWeekday, type = "l", main = "Weekday")
plot(stepsWeekend, type = "l", main = "Weekend")
```



